Symmetric Implicit Surface Microstructures for Hierarchical Design

### 2025-02-28 manuscript/main.tex
add TODO to modify domain of integration for triangles and tetrahedrons
rename unused latex files with .OLD extension
update latex files
push some edits

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ruiqi Chen}
\email{rchensix@alumni.stanford.edu}
\orcid{0002-5273-6277}
\author{Hardik Kabaria}
\email{hardikk@alumni.stanford.edu}
\author{Saigopal Nelaturi}
\email{snelaturi@carbon3d.com}
\affiliation{
  \institution{Carbon, Inc.}
  \streetaddress{1089 Mills Way}
  \city{Redwood City}
  \state{California}
  \country{USA}
  \postcode{94063}
}

\author{Adrian J. Lew}
\email{lewa@stanford.edu}
\affiliation{
  \institution{Stanford University}
  \streetaddress{440 Escondido Mall}
  \city{Stanford}
  \state{California}
  \country{USA}
  \postcode{94305}
}

@@

In short, this paper makes the following contributions:
% TODO(rchensix): Add more contributions
\begin{itemize}
  \item We develop a novel method to constrain the unit cell to intersect its bounding polygon/polyhedron at a right angle by formulating a constrained convex optimization problem that is always feasible.
\end{itemize}

@@

Implicit surfaces represent a geometric object through an equation of the form $\Phi(\mathbf{x}) = 0$ where $\Phi: \mathbb{R}^d \rightarrow \mathbb{R}$ is some function that evaluates a query point $\mathbf{x}$ and returns zero if $\mathbf{x}$ is on the surface, returns a negative value if $\mathbf{x}$ is inside the surface, or returns a positive value if $\mathbf{x}$ is outside the surface (note that this convention is not universal and other authors may use the opposite convention, i.e. points inside the surface have positive return values). The magnitude of the return value of the implicit surface function may have some geometric meaning, but this is not required. For example, signed distance functions are a subset of implicit surface equations where the magnitude of the return value denotes the distance from the query point $\mathbf{x}$ to the surface. In this work, we are interested in implicit surface functions that represent 2D and 3D geometries, so we assume that $d = 2$ or $d = 3$.

A symmetric implicit surface is defined in this work to be an implicit surface equation that has been constrained such that the geometry that it represents is unchanged after applying any transformation from a chosen symmetry group $\mathcal{S}$. In other words, $\Phi(\mathbf{x}) = \Phi(\mathbf{Q} \mathbf{x})$, where $\mathbf{Q}: \mathbb{R}^{d \times d} \in \mathcal{S}$ is an orthogonal matrix that represents a linear transformation of a point $\mathbf{x}$ about the origin. Note that for this work, we are restricting ourselves to linear transformations and excluding affine transformations. We want to constrain the implicit surface to match the symmetry group of the cell type in the unstructured scaffold mesh to tile: dihedral group of degree 3 and order 6 ($D_3$) for triangle meshes, dihedral group of degree 4 and order 8 ($D_4$) for quad meshes, achiral tetrahedral symmetry ($T_d$) for tetrahedron meshes, and achiral octahedral symmetry ($O_h$) for hexahedron meshes. The list of symmetry operators for these four symmetry groups is given in Appendix \ref{sec:symmetry_groups}. The necessity of symmetric implicit surfaces to tile the unstructured scaffold mesh is explained in detail in Section \ref{sec:tiling_mesh_scaffolds}, but in brief, it is the symmetry that ensures $C^0$ continuity at the boundaries of the hierarchical structure. Furthermore, symmetry is crucial for enabling the spatial transitions described in Section \ref{sec:spatial_transitions}.

To construct a symmetric implicit surface $\Phi(\mathbf{x})$, we first construct a basis function that is inherently symmetric. The symmetric basis function is constructed from some other conventional basis function $\psi(\mathbf{x}): \mathbb{R}^d \rightarrow \mathbb{R}$ by iterating over the symmetry operators $\mathbf{Q} \in \mathcal{S}$, feeding the transformed point $\mathbf{Qx}$ as input to $\psi$, and summing over all values of $\mathbf{Q}$:
%
\begin{equation}
  \label{eq:basis_function}
  \phi(\mathbf{x}) = K \sum\limits_{\mathbf{Q} \in \mathcal{S}} \psi(\mathbf{Qx})
\end{equation}
%
where $K$ is some normalization factor to make the basis function orthonormal. In this work, the $d$-dimensional Fourier series
%
\begin{equation}
  \label{eq:fourier_series}
  \psi_{\mathbf{f}}(\mathbf{x}) = e^{2\pi i \mathbf{f}^T \mathbf{x}}
\end{equation}
%
is used, where $\mathbf{f}: \mathbb{R}^d \in \mathcal{T}$ is an integer-valued vector that represents spatial frequencies in the Cartesian directions, and $\mathcal{T} \subset \mathbb{Z}^d$ is a subset of the $d$-dimensional integer lattice that is specific to the chosen scaffold mesh type. In other words, $\mathcal{T}$ contains all of the \textit{unique} spatial frequencies that may be used to construct a symmetric basis functions. Any spatial frequencies not in $\mathcal{T}$ do not need to be considered due to symmetry. For example, suppose the scaffold mesh is a quad mesh, then one representation for $\mathcal{T}$ is given by
%
\begin{equation}
  \mathcal{T} = \left\{ \begin{bmatrix} f_x \\ f_y \end{bmatrix} \in \mathbb{Z}^2 | f_y \geq 0 \: \text{and} \: f_x - f_y \geq 0 \right\}
\end{equation}
%
This representation of $\mathcal{T}$ for a quad mesh is depicted in Figure \ref{fig:unique_freqs_quad}(a).
%
\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{figures/quad_z_lattice_points.pdf}
  \caption{A subset $\mathcal{T}$ of the $\mathbb{Z}^2$ lattice defines unique spatial frequencies that can be used to generate unit cells with $D_4$ symmetry. Both (a) and (b) are valid representations for $\mathcal{T}$, and it is a matter of user preference to choose.}
  \label{fig:unique_freqs_quad}
\end{figure}
%
Note that if all of the symmetry operators in $S$ are applied to $\mathcal{T}$, then we would recover $\mathbb{Z}^2$. Also note that the representation of $\mathcal{T}$ is not unique; Figure \ref{fig:unique_freqs_quad}(b) shows an alternative representation that is equally valid. Representations of $\mathcal{T}$ used in this work for triangles, quadrilaterals, tetrahedrons, and hexahedrons are given in Appendix \ref{sec:symmetry_groups}. Additionally, note that by writing the Fourier series using complex exponentials (as expressed in Equation \ref{eq:fourier_series}), Equation \ref{eq:basis_function} is, in general, a function that takes in real-valued $d$-D points and returns complex valued scalars, i.e. $\phi_{\mathbf{f}}(\mathbf{x}): \mathbb{R}^d \rightarrow \mathbb{C}$. However, for the purpose of generating implicit surfaces, only the real or imaginary part (but not both) of the final expression for $\Phi(\mathbf{x})$ will be used.

There are many reasons, both mathematical and practical, as to why the Fourier series is an excellent choice for the foundational basis function, but the primary reasons are: Fourier series are an orthonormal basis, Fourier series have a very simple expression for the gradient (this is used in Section \ref{sec:face_normal_surfaces}); there are fast, robust algorithms available (most notably, the fast Fourier transform) for computing the cofficients of a Fourier expansion; and finally, several famous triply periodic minimal surfaces (TPMS) that are commonly used to construct hierarchical lattice structures (e.g. Schwarz primitive, gyroid) may be approximated using Fourier series \cite{wohlgemuth2001triply}.

Substituting Equation \ref{eq:fourier_series} into Equation \ref{eq:basis_function}, the symmetric basis function may be expressed as
%
\begin{equation}
  \label{eq:basis_function_actual}
  \phi_{\mathbf{f}}(\mathbf{x}) = K_{\mathbf{f}} \sum\limits_{\mathbf{Q} \in \mathcal{S}} e^{2 \pi i \mathbf{f}^T \mathbf{Q} \mathbf{x}}
\end{equation}
%
where $K_{\mathbf{f}}$ is the normalization coefficient given by
%
\begin{equation}
  \label{eq:normalization_coeff}
  K_{\mathbf{f}} = \frac{1}{\sqrt{ \iiint_\Omega \left( \sum\limits_{\mathbf{Q_1}, \mathbf{Q_2} \in \mathcal{S}} e^{2 \pi i \mathbf{f}^T (\mathbf{Q_1 - Q_2}) \mathbf{x}} \right) dV }}
\end{equation}
%
where the region of integration $\Omega$ is either the unit square (with side length of 1) centered at the origin for $d = 2$ or the unit cube (with side length of 1) centered at the origin for $d = 3$. In the integrand in the denominator of Equation \ref{eq:normalization_coeff}, due to the orthonormal property of Fourier series, only terms where $\mathbf{f}^T (\mathbf{Q_1 - Q_2}) = \mathbf{0}$ contribute to the integral.

@@

>>>>

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ruiqi Chen}
\email{rchensix@alumni.stanford.edu}
\orcid{0002-5273-6277}
\author{Hardik Kabaria}
\email{hardikk@alumni.stanford.edu}
\orcid{0000-0000-0000}
\author{Adrian J. Lew}
\email{lewa@stanford.edu}
\orcid{0000-0002-1559-7087}
\affiliation{
  \institution{Stanford University}
  \streetaddress{440 Escondido Mall}
  \city{Stanford}
  \state{California}
  \country{USA}
  \postcode{94305}
}

@@

In short, this paper makes the following contributions:
% TODO(rchensix): Add more contributions
\begin{itemize}
  \item We develop a novel method to construct hierarchical structures from an unstructured mesh scaffold comprised solely of triangles, quadrilaterals, tetrahedrons, or hexahedrons.
  \item Our method allows for multiple cell types to be prescribed, enabling spatially varying hierarchical structures.
  \item We enable constraining the unit cell to intersect its bounding polygon/polyhedron at a right angle by solving a constrained least squares problem.
\end{itemize}

@@

Implicit surfaces represent a geometric object through an equation of the form $\Phi(\mathbf{x}) = 0$ where $\Phi: \mathbb{R}^d \rightarrow \mathbb{R}$ is some function that evaluates a query point $\mathbf{x} \in \mathbb{R}^d$ and returns zero if $\mathbf{x}$ is on the surface, returns a negative value if $\mathbf{x}$ is inside the surface, or returns a positive value if $\mathbf{x}$ is outside the surface (note that this convention is not universal and some authors use the opposite convention). In this work, we are interested in implicit surface functions that represent 2D and 3D geometries, so we assume that $d = 2$ or $d = 3$. The magnitude of the return value of the implicit surface function may have some geometric meaning, but this is not required. For example, signed distance functions (SDFs) are a subset of implicit surface equations where the magnitude of the return value denotes the distance from $\mathbf{x}$ to the surface.

A symmetric implicit surface is defined in this work to be an implicit surface equation that is invariant under transformations from a chosen point group $\mathcal{S}$. In other words, $\Phi(\mathbf{x}) = \Phi(\mathbf{Q} \mathbf{x})$, where $\mathbf{Q}: \mathbb{R}^{d \times d} \in \mathcal{S}$ is an orthonormal matrix that represents an isometry (i.e. rotation, reflection, inversion, or rotoinversion of a point $\mathbf{x}$ about the origin). To construct an implicit surface suitable for generating hierarhical structures, we want to constrain the implicit surface to match the symmetry group of the cell type in the unstructured scaffold mesh to tile: dihedral group of degree 3 and order 6 ($D_3$) for triangle meshes, dihedral group of degree 4 and order 8 ($D_4$) for quadrilateral meshes, achiral tetrahedral symmetry ($T_d$) for tetrahedron meshes, and achiral octahedral symmetry ($O_h$) for hexahedron meshes. The list of symmetry operators for these four symmetry groups is given in Appendix \ref{sec:symmetry_groups}. The necessity of symmetric implicit surfaces to tile the unstructured scaffold mesh is explained in detail in Section \ref{sec:tiling_mesh_scaffolds}, but in brief, it is the symmetry that ensures $C^0$ continuity at the boundaries of the hierarchical structure. Furthermore, symmetry is crucial for enabling the spatial transitions described in Section \ref{sec:spatial_transitions}.

To construct a symmetric implicit surface $\Phi(\mathbf{x})$, we first construct an orthonormal basis function $\phi(\mathbf{x}): \mathbb{R}^d \rightarrow \mathbb{R}$ that is inherently symmetric. The symmetric basis function is constructed from some other conventional orthonormal basis function $\psi(\mathbf{x}): \mathbb{R}^d \rightarrow \mathbb{R}$ by iterating over the symmetry operators $\mathbf{Q}$ and summing together $\psi$ composed with the linear transformation of $\mathbf{x}$ by $\mathbf{Q}$:
%
\begin{equation}
  \label{eq:basis_function}
  \phi(\mathbf{x}) = K \sum\limits_{\mathbf{Q} \in \mathcal{S}} \psi(\mathbf{Qx})
\end{equation}
%
where $K$ is some normalization factor to make the basis function orthonormal. In this work, the $d$-dimensional Fourier series
%
\begin{equation}
  \label{eq:fourier_series}
  \psi_{\mathbf{f}}(\mathbf{x}) = e^{2\pi i \mathbf{f}^T \mathbf{x}}
\end{equation}
%
is used, where $\mathbf{f}: \mathbb{R}^d \in \mathcal{T}$ is an integer-valued vector that represents spatial frequencies in the Cartesian directions, and $\mathcal{T} \subset \mathbb{Z}^d$ is a subset of the $d$-dimensional integer lattice that is specific to the chosen scaffold mesh type. In other words, $\mathcal{T}$ contains all of the \textit{unique} spatial frequencies that may be used to construct a symmetric basis functions. Any spatial frequencies not in $\mathcal{T}$ do not need to be considered due to symmetry. For example, suppose the scaffold mesh is a quadrilateral mesh, then one representation for $\mathcal{T}$ is given by
%
\begin{equation}
  \mathcal{T} = \left\{ (f_x, f_y) \in \mathbb{Z}^2 | f_y \geq 0 \: \text{and} \: f_x - f_y \geq 0 \right\}
\end{equation}
%
This representation of $\mathcal{T}$ for a quadrilateral mesh is depicted in Figure \ref{fig:unique_freqs_quad}(a).
%
\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{figures/quad_z_lattice_points.pdf}
  \caption{A subset $\mathcal{T}$ of the $\mathbb{Z}^2$ lattice defines unique spatial frequencies that can be used to generate unit cells with $D_4$ symmetry. Both (a) and (b) are valid representations for $\mathcal{T}$, and it is a matter of user preference to choose.}
  \label{fig:unique_freqs_quad}
\end{figure}
%
Note that if all of the symmetry operators in $S$ are applied to $\mathcal{T}$, then we would recover $\mathbb{Z}^2$. Also note that the representation of $\mathcal{T}$ is not unique; Figure \ref{fig:unique_freqs_quad}(b) shows an alternative representation that is equally valid. Representations of $\mathcal{T}$ used in this work for triangles, quadrilaterals, tetrahedrons, and hexahedrons are given in Appendix \ref{sec:symmetry_groups}. Additionally, note that by writing the Fourier series using complex exponentials (as expressed in Equation \ref{eq:fourier_series}), Equation \ref{eq:basis_function} is, in general, a function that takes in real-valued $d$-D points and returns complex valued scalars, i.e. $\phi_{\mathbf{f}}(\mathbf{x}): \mathbb{R}^d \rightarrow \mathbb{C}$. However, for the purpose of generating implicit surfaces, only the real or imaginary part (but not both) of the final expression for $\Phi(\mathbf{x})$ will be used.

There are many reasons, both mathematical and practical, as to why the Fourier series is an excellent choice for the foundational basis function, but the primary reasons are: Fourier series are an orthonormal basis, Fourier series have a very simple expression for the gradient (this is used in Section \ref{sec:face_normal_surfaces}); there are fast, robust algorithms available (most notably, the fast Fourier transform) for computing the cofficients of a Fourier expansion; and finally, several famous triply periodic minimal surfaces (TPMS) commonly used to construct hierarchical lattice structures (e.g. Schwarz primitive, gyroid) may be approximated using Fourier series \cite{wohlgemuth2001triply}.

Substituting Equation \ref{eq:fourier_series} into Equation \ref{eq:basis_function}, the symmetric basis function may be expressed as
%
\begin{equation}
  \label{eq:basis_function_actual}
  \phi_{\mathbf{f}}(\mathbf{x}) = K_{\mathbf{f}} \sum\limits_{\mathbf{Q} \in \mathcal{S}} e^{2 \pi i \mathbf{f}^T \mathbf{Q} \mathbf{x}}
\end{equation}
%
where $K_{\mathbf{f}}$ is the normalization coefficient given by
%
\begin{equation}
  \label{eq:normalization_coeff}
  K_{\mathbf{f}} = \frac{1}{\sqrt{ \iiint_\Omega \left( \sum\limits_{\mathbf{Q_1}, \mathbf{Q_2} \in \mathcal{S}} e^{2 \pi i \mathbf{f}^T (\mathbf{Q_1 - Q_2}) \mathbf{x}} \right) dV }}
\end{equation}
%
% TODO(rchensix): Need to explain "extended" domain, i.e. unit cube for tetrahedrons and unit rhombus for triangles, which permit orthogonal Fourier series basis functions.

where the region of integration $\Omega$ is either the unit square (with side length of 1) centered at the origin for quadrilateral  or the unit cube (with side length of 1) centered at the origin for $d = 3$. In the integrand in the denominator of Equation \ref{eq:normalization_coeff}, due to the orthonormal property of Fourier series, only terms where $\mathbf{f}^T (\mathbf{Q_1 - Q_2}) = \mathbf{0}$ contribute to the integral.

@@

Dilating, eroding, and inverting the implicit surface unit cell is trivial.
